Distributed Adaptive Norm Estimation For Blind System\\Identification in Wireless Sensor Networks

### 2022-10-25 paper.tex
Jesper's comments

Distributed algorithms have been an active area of research for quite some time, with numerous control, optimization, and signal processing applications.
With the ever-growing number of smart multimedia devices in today's surroundings providing ubiquitous processing and communication capabilities, distributed audio and speech signal processing have also found their way into the spotlight.
Algorithms for distributed signal estimation \cite{5483092}, noise control and echo cancellation \cite{9670697}, as well as beamforming \cite{6663655,6329934,MARKOVICHGOLAN20154,6309434} to name a few, have been proposed.
Another important task in audio and communication applications, which however has not recived as much attention, is multi-channel system identification, i.e. estimating channel responses in the time or frequency domain.
Distributed single-input-multiple-output (SIMO) blind system identification (BSI) has been addressed with adaptive cross-relation-based (CR) algorithms \cite{yuDistributedBlindSystem2014, liuDistributedBlindIdentification2016}.
In this context, we recently introduced an adaptive CR-based algorithm \cite{blochbergerDBSI} using the alternating direction method of multipliers (ADMM) \cite{boydDistributedOptimizationStatistical2011}.

All of the aforementioned distributed BSI algorithms rely on shared information between neighboring sensor nodes within the network.
However, the CR-based BSI task necessitates a non-triviality constraint on the full system to be identified (we refer the reader to, e.g., \cite{huangAdaptiveMultichannelLeast2002,huangClassFrequencydomainAdaptive2003}), which manifests itself as one or more variables that require network-wide shared information for their computation.
The algorithm in \cite{blochbergerDBSI} relies on node-wise values being relayed throughout the network.
In contrast, to overcome the need for the network to be fully connected, \cite{yuDistributedBlindSystem2014, liuDistributedBlindIdentification2016} use an average consensus \cite{xiaoFastLinearIterations2004} approach where a secondary recursion estimates the global variable for each signal frame.
Both approaches introduce additional transmissions of variables between nodes, the number of which, depending on network and neighborhood size, can be substantial or even unfeasable.

In this paper, we extend our algorithm in \cite{blochbergerDBSI} with a distributed averaging-based \cite{xiaoFastLinearIterations2004} estimation scheme for the global variable.
This approach allows us, similarly to \cite{6334305,9914798}, to compute the global variable without the need of a fully connected network or broadcasting.
We also introduce a mixing factor to include instantaneous values into the averaging recursion, which then allows us (i) to reduce the number of secondary iterations significantly (from 50 in existing approaches, down to 1) and (ii) track time-varying systems.
Simulation results demonstrate how the extension leads to BSI performance close to an optimal case where all information is available without the need of a fully connected network or a large number of estimation iterations.

\section{DISTRIBUTED ADAPTIVE BSI IN SENSOR NETWORKS WITH ONLINE-ADMM}
\label{sec:dbsi}
In this section, we briefly outline the adaptive SIMO BSI algorithm proposed in \cite{blochbergerDBSI}, setting the scene for the distributed averaging scheme proposed herein.
We refer the reader to the publication cited above for the full derivation of the algorithm.

\subsection[]{Sensor network}
We assume a set of nodes with one sensor each, that is, each node provides one signal.
Let this set of nodes be \(\Mset \triangleq \{1,\ldots,M\}\) and \(\mathcal{E}\) be a set of edge tuples, which connect the nodes forming a sensor network.
Each edge is an unordered pair of node indices \(\{i,j\} \in \mathcal{E}\), which represents the communication link between them.
We allow \(\{i,i\} \in \mathcal{E}\), in order to simplify notation later, however, this does not represent a link but rather indicates the obvious fact that node \(i\) has information about itself.
Furthermore, let \(\Nset_i = \{j|\{i,j\} \in \mathcal{E}\}\) denote the neighborhood of node \(i\).
We can define the symmetric adjacency matrix \(\mtxb{C}\) with elements \(C_{ij} = 1\) if \(\{i,j\} \in \mathcal{E}\) and 0 otherwise.
It may be noted that in \cite{blochbergerDBSI}, the pairs \(\{i,j\} \in \mathcal{E}\) are ordered, which yields a directed graph and leads to two sets of neighborhoods (``transmit'' and ``receive'') and therefore a non-symmetric adjacency matrix.
For the sake of brevity, we limit the explanations in this paper to the undirected case.
The case of directed graphs can be extended directly from the presented results.

@@

\subsection[]{Distributed BSI with Online-ADMM}
In BSI, the cross-relation problem formulation only uses output signals \(\x_i(n)\), exploiting relative information between them, to identify the system, i.e., the acoustic or communication channels \(\h = [\h_1^\T,\,...,\,\h_M^\T]^\T\).
The solution to this problem is found by the minimization problem \cite{langtongBlindIdentificationEqualization1994,huangAdaptiveMultichannelLeast2002,huangClassFrequencydomainAdaptive2003,blochbergerDBSI}:
\begin{equation}
    \begin{aligned}
        \hat{\h} = \arg \min_{\h} \quad &\h^\herm \R \h \\
        \text{s.t. } \quad &\|\h\| = 1,
    \end{aligned}\label{eq:frequency_domain:min_prob}
\end{equation}
where \(\hat{\h} = [\hat{\h}_1^\T,\,...,\,\hat{\h}_M^\T]^\T\) is the vector of estimated channel responses and \(\R\) is the so-called cross-relation (CR) matrix.
In \cite{blochbergerDBSI}, this problem is solved by a distributed adaptive algorithm that is based on separating the problem \eqref{eq:frequency_domain:min_prob} into node-wise subproblems,
\begin{equation}
    \begin{aligned}
        \underset{\w,\,\h}{\min} \quad &\sum_{i \in \Mset} \w_i^\herm \aRhof_i \w_i\\
        \text{s.t.} \quad &(\w_i)_j = \h_{\mathcal{G}(i,j)}\quad i \in \Mset,\,j \in \Nset_i\\
        &\|\h\| = 1.
    \end{aligned}\label{eq:general_consensus_admm:min_prob}
\end{equation}
Here, each subproblem is represented by a cost function \(\w_i^\herm \aRhof_i \w_i\), with the node-local channel estimates \(\w_i\) and CR matrix \(\aRhof_i\).
Both are analogous to \(\h_i\) and \(\R\) in \eqref{eq:frequency_domain:min_prob}, however, only using a subset of channels, corresponding to \(\Nset_i \subset \Mset\).
The first constraint enforces a consensus of local estimates, and \(\mathcal{G}(i,j)\) is a map equating local estimates of nodes connected by edge \((i,j) \in \mathcal{E}\).
The general-form consensus alternating direction method of multipliers (ADMM) \cite{boydDistributedOptimizationStatistical2011} is applied in an adaptive updating scheme (Online-ADMM) \cite{wangOnlineAlternatingDirection2013,hosseiniOnlineDistributedADMM2014}.
ADMM leads to update steps for local, consensus, and dual variables.
We refer the reader to \cite{blochbergerDBSI} for details.
The algorithm's consensus variable is the system's estimate, \(\hat{\h}\).
Each node \(i\) computes the respective subvector \(\hat{\h}_i\) locally, which is the computation step of interest for this extension.
It is defined as
\begin{equation}
    \hat{\h_i}^{(n)} = \frac{\bar{\h}_i^{(n)}}{\sqrt{\sum_{j \in \Mset} \|\bar{\h}_j^{(n)}\|^2}}\label{eq:online_admm:consensus_update}
\end{equation}
for each node \(i \in \Mset\) with the local unnormalized consensus \(\bar{\h}_i^{(n)}\), a combination of neigborhood averages of local estimates and dual variables, see \cite{blochbergerDBSI}.
The computation of the denominator of \eqref{eq:online_admm:consensus_update} requires  \(\|\bar{\h}_j^{(n)}\|\) from all nodes \(j \in \Mset\), which however, is not possible without a fully-connected network or broadcasting.
In \cite{blochbergerDBSI}, we assume that partial squared norms are relayed throughout the network until all nodes have the information necessary.

In \cite{yuDistributedBlindSystem2014,liuDistributedBlindIdentification2016},
a global value is estimated by a related iterative averaging approach, combining values within neighborhoods until convergence to a network-wide average.

Subsequently, we introduce an extension to the algorithm described in \cite{blochbergerDBSI} using a fastest distributed linear averaging (FDLA) approach \cite{xiaoFastLinearIterations2004} to avoid the need of network-wide data transmission.
The approach described in \cite{yuDistributedBlindSystem2014,liuDistributedBlindIdentification2016} employs a related iterative averaging scheme within a distributed gradient descent algorithm for BSI in order to estimate a global variable, however, necessitating many iterations per frame.
To alleviate this, we further introduce an adaptive mixing factor to include instantaneous values in the recursion to separate avaraging iterations onto time frames of the adaptive algorithm.

@@

    \caption[]{Median NPM of 30 Monte-Carlo runs over time for optimal case where all node information is available (fully connected network/broadcasting), fixed values of \(\gamma_i\) with \(K \in \{1,10\}\), adaptive \(\gamma_i\) with \(K=1\). A moving average filter was applied to result curves for better readability (compare transparent and opaque lines). Hatched area indicates frames used in computations for \autoref{fig:simulations:avgNPMgamma}.}

@@

Given that the BSI algorithm being extended \cite{blochbergerDBSI} is adaptive and the unnormalized consensus estimates \(\bar{\h}_i^{(n)}\) are therefore time-varying, computing a distributed sum of their norms has to be performed adaptively.
The baseline approach is to compute \(K\) iterations \eqref{eq:adaptivenormest:distlincomb} per frame index \(n\).
For the baseline iterative approach to converge, \(K\) has to be sufficiently large (e.g., \(K=50\) in \cite{yuDistributedBlindSystem2014,liuDistributedBlindIdentification2016}).
This is suboptimal in terms of communication cost, as each iteration requires that a node \(i\) shares the averaging variable \(\phi_i(k)\) with its neighborhood \(\Nset_i\).

@@

For each \(n\), we could set the initial values \(\phi_i^{(n)}(0)\) to the result of the \(K\)-th iteration \(\phi_i^{(n-1)}(K)\) of the previous frame \(n-1\).

@@

Therefore, we factor in the instantaneous approximation, the weighted in-neighborhood sum \(\eta_i^{(n)} = \sum_{j \in \Nset_i} W_{ij} \|\bar{\h}_i^{(n)}\|^2\), which will introduce new information gained by more accurate estimates of \(\bar{\h}_i^{(n)}\).
We add this information by linearly combining it with the \(K\)-th iterate of \(\phi_i^{(n)}(k)\) and set it as the initial value for the next frame:

@@

where \(\gamma_i\) is a mixing factor.

@@

If the absolute difference between subsequent frames \(| \eta_i^{(n)} - \eta_i^{(n-1)} | \to 0\), then the estimation of \(\bar{\h}_i\) is converging to a steady state.
The algorithm's emphasis should then lie on norm estimation, i.e., the distributed averaging recursion, which means \(\gamma_i \to 0\).
We heuristically set \(\gamma_i^{(n)}\) proportional to the absolute difference of instantaneous approximations between subsequent frames and set an upper limit at 1,

@@

Refer to \autoref{alg:davg_norm_est} for the order of computations of the extended algorithm we introduced here.

\section{Evaluation}
\label{sec:simulations}
\begin{figure}[t]
    \centering
    \input{simulations/plots/icassp2023-static/icassp2023-static.pgf}
    \vspace*{-0.6cm}
    \caption[]{Mean of 500 post-convergence frames of median NPM of 30 Monte-Carlo runs for fixed \(\gamma_i\) on the interval \([0.0, 0.04]\) and \(K \in \{1,2,10\}\). Results for optimal algorithm, and adaptive \(\gamma_i\) included for comparison.}
    \label{fig:simulations:avgNPMgamma}
\end{figure}
\subsection[]{Communication cost}
\label{sec:transcost}
To describe the communication cost within the network, we count the number of variables transmitted per time frame \(n\).
The most useful measure in the case of this work is to analyze the cost per node.
We compare the following communication schemes within the context of the algorithm \cite{blochbergerDBSI} this paper proposes the extension to:
\begin{itemize}
    \itemsep-0.2em
    \item[(1)] A fully-connected network, where node \(i\) communicates \(\|\bar{\h}_i\|^2\) to all other nodes \(\{j \in \Nset_i = \Mset\}\) directly, i.e., the neighborhood is the full network.
    \item[(2)] The node \(i\) communicates \(\|\bar{\h}_i\|^2\) only to neighboring nodes \(\{j \in \Nset_i \subset \Mset\}\) and applies \(K\) distributed averaging iterations.
\end{itemize}
\autoref{tab:transcost:table} compares the number of transmit and receive operations node \(i\) has to apply in order to compute \eqref{eq:online_admm:consensus_update}.
Further, it shows the complexity of the problem node \(i\) has to solve according to \cite{blochbergerDBSI} given the two types of network topology, assuming it uses all information it receives.
The advantage of the fully distributed algorithm becomes clear when \(M \gg N_i\), as the additional communication and local complexity depends on neighborhood size \(N_i\) and not on network size \(M\).
\vspace*{-0.6em}
\renewcommand{\arraystretch}{1.2}
\begin{table}[h]
    \centering
    \begin{tabular}{ |l|l|l|l| }
        \hline
        & Trans. Ops. & Rec. Ops. & Local Compl. \\
        \hline\hline
        (1) Fully con. & \(M-1\) & \(M-1\) & \(\mathcal{O}(M^2 L^2)\) \\
        \hline
        (2) Neighborh. & \(N_i K\) & \(N_i K\) & \(\mathcal{O}(N_i^2 L^2)\)\\ 
        \hline
    \end{tabular}
    \caption[]{Communication cost per node and frame \(n\) for computation of \eqref{eq:online_admm:consensus_update}.}
    \label{tab:transcost:table}
\end{table}
\renewcommand{\arraystretch}{1.0}

\subsection[]{Simulations}
To evaluate the effectiveness of the proposed extension, simulations were run, where we define the error measure as the normalized projection misalignment
\begin{equation}
    \begin{aligned}
        \text{NPM}(n) &= 20\,\log_{10} \frac{\left\| \mtxb{e} \right\|}{\left\|\hat{\h}^{(n)}\right\|},\\
        \text{with}\quad \mtxb{e} &= \hat{\h}^{(n)} - \frac{\h^\T \hat{\h}^{(n)}}{\h^\T \h}\hat{\h}^{(n)},
    \end{aligned}
\end{equation}
commonly used in BSI to compare the estimated \(\hat{\h}^{(n)}\) and true \(\h\).

The first simulation setup is a network with \(M=5\) nodes arranged in a ring topology, each node having \(N_i=3\) neighbors (including node \(i\)).
The input signal is zero-mean white Gaussian noise (WGN), the impulse responses of length \(L=16\) are drawn from a normal distribution, and at each channel, independent WGN is added with \(\text{SNR}=10\,\text{dB}\).
The norms of the impulse responses are scaled to random values drawn from the uniform distribution \(\mathcal{U}_{[0.5,2.0]}\).
A comparison is made between the following cases:
\begin{itemize}
    \itemsep-0.2em
    \item[(a)] \(\|\bar{\h}_i\|^2\) of all \(i \in \Mset\) is available for global norm computation (``optimal'' in \autoref{fig:simulations:NPMtime} \& \autoref{fig:simulations:avgNPMgamma}),
    \item[(b)] Inter-neighborhood communication with fixed \(\gamma_i \in [0.0, 0.4]\) and \(K \in \{1,2,10\}\),
    \item[(c)] Inter-neighborhood communication with adaptive \(\gamma_i\) \eqref{eq:adaptivenormest:adaptivegamma} and \(K=1\).
\end{itemize}
\begin{algorithm}[t]
    \caption{ADMM BSI with distributed-averaging-based adaptive estimation of norm values}\label{alg:davg_norm_est}
    \(\W \gets\) \eqref{eq:adaptivenormest:fdlaminprob}\;
    \(\eta_i^{(n-1)} \gets 1, \forall i \in \Mset\)\;
    \For(){\(n=0\dots\)}
    {
        \For(){\(i \in \Mset\)}
        {
            \emph{The steps before are as introduced in }\cite{blochbergerDBSI}\\
            \dotfill\\
            \(\eta_i^{(n)} \gets \sum_{j \in \Nset_i} W_{ij} \|\bar{\h}_i^{(n)}\|^2\)\;
            \(\gamma_i^{(n)} \gets \min \left\lbrace \frac{| \eta_i^{(n)} - \eta_i^{(n-1)} |}{\eta_i^{(n-1)}},\,1\right\rbrace\)\;
            \eIf(){\(n = 0\)}
            {
                \(\phi_i^{(0)}(0) \gets \|\bar{\h}_i^{(0)}\|^2\)\;
            }
            {
                \(\phi_i^{(n)}(0) \gets \gamma_i^{(n)} \eta_i^{(n)} + (1-\gamma_i^{(n)}) \phi_i^{(n-1)}(K)\)\;
            }
            \For(){\(k=0,\dots,K\)}
            {
                Transmit \(\phi_i^{(n)}(k)\) to nodes  \(j \in \Nset_i\)\;
                \(\phi_i^{(n)}({k+1}) \gets \sum_{j \in \Nset_i} W_{ij} \phi_j^{(n)}({k})\)\;
            }
            \(\hat{\h_i}^{(n)} \gets \frac{\bar{\h}_i^{(n)}}{\sqrt{M \phi_i^{(n)}(K)}}\)\;
            \dotfill\\
            \emph{The steps after are as introduced in }\cite{blochbergerDBSI}\\
        }
    }
\end{algorithm}
\autoref{fig:simulations:NPMtime} shows the median of 30 Monte-Carlo runs.
We can observe that when fixing \(\gamma_i\), a tradeoff between convergence speed and steady-state error arises at low iteration counts (cf. also \autoref{fig:simulations:avgNPMgamma}).
With \(K=10\), the performance comes close to the optimal case at the cost of additional communication between nodes.
This in contrast to the case with adaptive \(\gamma_i\) where even at \(K=1\), i.e., no additional communication for the recursive estimation scheme, convergence speed and steady-state error are close to the optimal case.

>>>>

Distributed algorithms have been an active area of research for quite some time, with numerous control, optimization, and signal processing applications.
With the ever-growing number of smart multimedia devices in today's surroundings providing ubiquitous processing and communication capabilities, distributed audio and speech signal processing have also found their way into the spotlight.
Algorithms for distributed signal estimation \cite{5483092}, noise control and echo cancellation \cite{9670697}, as well as beamforming \cite{6663655,6329934,MARKOVICHGOLAN20154,6309434} to name a few, have been proposed.
Another essential task in audio and communication applications, which has received less attention, is multi-channel system identification, i.e., estimating channel responses in the time or frequency domain.
Distributed single-input-multiple-output (SIMO) blind system identification (BSI) has been addressed with adaptive cross-relation-based (CR) algorithms \cite{yuDistributedBlindSystem2014, liuDistributedBlindIdentification2016}.
In this context, we recently introduced an adaptive CR-based algorithm \cite{blochbergerDBSI} using the alternating direction method of multipliers (ADMM) \cite{boydDistributedOptimizationStatistical2011}.

The aforementioned distributed BSI algorithms rely on shared information between neighboring sensor nodes within the network.
However, the CR-based BSI task necessitates a non-triviality constraint on the full system to be identified (we refer the reader to, e.g., \cite{huangAdaptiveMultichannelLeast2002,huangClassFrequencydomainAdaptive2003} for details), which manifests itself as one or more variables that require network-wide shared information for their computation.
The algorithm in \cite{blochbergerDBSI} relies on node-wise values being relayed throughout the network.
In contrast, to overcome the need for the network to be fully connected, \cite{yuDistributedBlindSystem2014, liuDistributedBlindIdentification2016} use an average consensus \cite{xiaoFastLinearIterations2004} approach where a secondary recursion estimates the global variable for each signal frame.
Both approaches introduce additional transmissions of variables between nodes, the number of which, depending on network and neighborhood size, can be substantial or even unfeasible depending on the application.

In this paper, we extend our algorithm in \cite{blochbergerDBSI} with a distributed averaging-based \cite{xiaoFastLinearIterations2004} estimation scheme for the global variable.
This approach allows us, similarly to \cite{6334305,9914798}, to compute the global variable without needing a fully connected network or broadcasting.
We also introduce a mixing factor to include instantaneous values into the averaging recursion, which then allows us (i) to reduce the number of secondary iterations significantly (from 50 in existing approaches down to 1) and (ii) track time-varying systems.
Simulation results demonstrate how the proposed reduced-communication algorithm using in-neighborhood information delivers BSI performance close to an idealized one where network-wide information is available.

\section{DISTRIBUTED ADAPTIVE BSI IN SENSOR NETWORKS WITH ONLINE-ADMM}
\label{sec:dbsi}
In this section, we briefly outline the adaptive SIMO BSI algorithm proposed in \cite{blochbergerDBSI}, setting the scene for the distributed averaging scheme proposed herein.
We refer the reader to \cite{blochbergerDBSI} for the full derivation of the algorithm.

\subsection[]{Sensor network}
We assume a set of network nodes with one sensor each, that is, each node provides one signal.
Let this set of nodes be \(\Mset \triangleq \{1,\ldots,M\}\) and \(\mathcal{E}\) be a set of edge tuples, which connect the nodes forming a sensor network.
Each edge is an unordered pair of node indices \(\{i,j\} \in \mathcal{E}\), which represents the communication link, which we assume is instantaneoud and error free, between them.
We allow \(\{i,i\} \in \mathcal{E}\), in order to simplify notation later, however, this does not represent a link but rather indicates the obvious fact that node \(i\) has information about itself.
Furthermore, let \(\Nset_i = \{j|\{i,j\} \in \mathcal{E}\}\) denote the neighborhood of node \(i\).
We can define the symmetric adjacency matrix \(\mtxb{C}\) with elements \(C_{ij} = 1\) if \(\{i,j\} \in \mathcal{E}\) and 0 otherwise.
It may be noted that in \cite{blochbergerDBSI}, the pairs \(\{i,j\} \in \mathcal{E}\) are ordered, which yields a directed graph and leads to two sets of neighborhoods (``transmit'' and ``receive'') and therefore a non-symmetric adjacency matrix.
For the sake of brevity, we limit the explanations in this paper to the undirected case.
The case of directed graphs can be extended directly from the presented results.

@@

\subsection[]{Distributed BSI with Online-ADMM}
In BSI, the cross-relation problem formulation only uses output signals \(\x_i(n)\), exploiting relative information between them, to identify the system, i.e., the acoustic or communication channels \(\h = [\h_1^\T,\,...,\,\h_M^\T]^\T\).
The solution to this problem is found by the minimization problem \cite{langtongBlindIdentificationEqualization1994,huangAdaptiveMultichannelLeast2002,huangClassFrequencydomainAdaptive2003,blochbergerDBSI}:
\begin{equation}
    \begin{aligned}
        \hat{\h} = \arg \min_{\h} \quad &\h^\herm \R \h \\
        \text{s.t. } \quad &\|\h\| = 1,
    \end{aligned}\label{eq:frequency_domain:min_prob}
\end{equation}
where \(\hat{\h} = [\hat{\h}_1^\T,\,...,\,\hat{\h}_M^\T]^\T\) is the vector of estimated channel responses and \(\R\) is the so-called cross-relation (CR) matrix.
In \cite{blochbergerDBSI}, this problem is solved by a distributed adaptive algorithm that is based on separating the problem \eqref{eq:frequency_domain:min_prob} into node-wise subproblems,
\begin{equation}
    \begin{aligned}
        \underset{\w,\,\h}{\min} \quad &\sum_{i \in \Mset} \w_i^\herm \aRhof_i \w_i\\
        \text{s.t.} \quad &(\w_i)_j = \h_{\mathcal{G}(i,j)}\quad i \in \Mset,\,j \in \Nset_i\\
        &\|\h\| = 1.
    \end{aligned}\label{eq:general_consensus_admm:min_prob}
\end{equation}
Here, each subproblem is represented by a cost function \(\w_i^\herm \aRhof_i \w_i\), with the node-local channel estimates \(\w_i\) and CR matrix \(\aRhof_i\).
Both are analogous to \(\h_i\) and \(\R\) in \eqref{eq:frequency_domain:min_prob}, however, only using a subset of channels, corresponding to \(\Nset_i \subset \Mset\).
The first constraint enforces a consensus of local estimates, and \(\mathcal{G}(i,j)\) is a map equating local estimates of nodes connected by edge \((i,j) \in \mathcal{E}\).
The general-form consensus alternating direction method of multipliers (ADMM) \cite{boydDistributedOptimizationStatistical2011} is applied in an adaptive updating scheme (Online-ADMM) \cite{wangOnlineAlternatingDirection2013,hosseiniOnlineDistributedADMM2014}.
ADMM leads to update steps for local, consensus, and dual variables.
We refer the reader to \cite{blochbergerDBSI} for details.
The algorithm's consensus variable is the system's estimate, \(\hat{\h}\).
Each node \(i\) computes the respective subvector \(\hat{\h}_i\) locally, which is the computation step of interest for the extension proposed in this paper.
It is defined as
\begin{equation}
    \hat{\h}_i^{(n)} = \frac{\bar{\h}_i^{(n)}}{\sqrt{\sum_{j \in \Mset} \|\bar{\h}_j^{(n)}\|^2}}\label{eq:online_admm:consensus_update}
\end{equation}
for each node \(i \in \Mset\) with the local unnormalized consensus \(\bar{\h}_i^{(n)} = \bar{\w}_i^{(n)} + \frac{1}{\rho} \bar{\uu}_i^{(n-1)}\), a combination of in-neigborhood averages of node-wise local estimates \(\bar{\w}_i^{(n)}\) and dual (Lagrangian) variables \(\bar{\uu}_i^{(n-1)}\), where \(\rho\) is the ADMM penalty parameter.
A proper introduction of these variables is unfortunately out of scope for this paper, so for details, see \cite{blochbergerDBSI}.
The computation of the denominator of \eqref{eq:online_admm:consensus_update} requires  \(\|\bar{\h}_j^{(n)}\|\) from all nodes \(j \in \Mset\), which however, is not possible without a fully-connected network or broadcasting.
In \cite{blochbergerDBSI}, we assume that partial squared norms are relayed/broadcasted throughout the network until all nodes have the information necessary.

Subsequently, we introduce an extension to the algorithm described in \cite{blochbergerDBSI} using a fastest distributed linear averaging (FDLA) approach \cite{xiaoFastLinearIterations2004} to avoid the need for network-wide data transmission.
A related approach described in \cite{yuDistributedBlindSystem2014,liuDistributedBlindIdentification2016} employs a similar iterative averaging scheme within a distributed gradient descent algorithm for BSI in order to estimate a global variable.
It, however, necessitates many iterations per frame.
To alleviate this, we further introduce an adaptive mixing factor to include instantaneous values in the recursion to separate averaging iterations onto time frames of the adaptive algorithm.

@@

    \caption[]{Median NPM of 30 Monte-Carlo runs over time for ideal case where all node information is available (fully connected network/broadcasting), fixed values of \(\gamma_i\) with \(K \in \{1,10\}\), adaptive \(\gamma_i\) with \(K=1\). A moving average filter was applied to result curves for better readability (compare transparent and opaque lines). Hatched area indicates frames used in computations for \autoref{fig:simulations:avgNPMgamma}.}

@@

Given that the unnormalized consensus estimates \(\bar{\h}_i^{(n)}\) \cite{blochbergerDBSI} are time-varying, computing a distributed sum of their norms has to be performed adaptively as well.
The baseline approach is to compute \(K\) iterations \eqref{eq:adaptivenormest:distlincomb} per frame index \(n\), where \(K\) has to be sufficiently large for convergence (e.g., \(K=50\) in \cite{yuDistributedBlindSystem2014,liuDistributedBlindIdentification2016}).
This is undesirable, as each iteration requires that a node \(i\) shares the averaging variable \(\phi_i(k)\) with its neighborhood \(\Nset_i\).

@@

One obvious approach would be - at each frame \(n\) - to set the initial values \(\phi_i^{(n)}(0)\) to the result of the \(K\)-th iteration \(\phi_i^{(n-1)}(K)\) of the previous frame \(n-1\).

@@

Instead, we factor in the instantaneous approximation, the weighted in-neighborhood sum \(\eta_i^{(n)} = \sum_{j \in \Nset_i} W_{ij} \|\bar{\h}_i^{(n)}\|^2\), which will introduce new information gained by more accurate estimates of \(\bar{\h}_i^{(n)}\).
We add this information by linearly combining it with the \(K\)-th iterate of \(\phi_i^{(n-1)}(k)\) and set it as the initial value fo the following frame

@@

where \(0 \leq \gamma_i \leq 1\) is a mixing factor.

@@

Experimental observations show that convergence is faster with larger \(\gamma_i\), while with smaller, steady-state error is lower.
We postulate that when the estimation of \(\bar{\h}_i\) is converging to a steady state, the absolute difference between subsequent frames \(| \eta_i^{(n)} - \eta_i^{(n-1)} | \to 0\).
Then - after faster convergence with larger \(\gamma_i\) - the algorithm's emphasis should lie on norm estimation, i.e., the distributed averaging recursion, which means \(\gamma_i \to 0\).
Based on this heuristic, we set \(\gamma_i^{(n)}\) proportional to the absolute difference of instantaneous approximations between subsequent frames and set an upper limit at 1,

@@

The adaptive estimate \(\phi_i^{(n)}(K)\) is then used in the denominator of \eqref{eq:online_admm:consensus_update} as \(\sqrt{M \, \phi_i^{(n)}(K)}\).
Refer to \autoref{alg:davg_norm_est} for the order of computations of the extended algorithm we introduced here.

\section{Evaluation}
\label{sec:simulations}
\begin{figure}[t]
    \centering
    \input{simulations/plots/icassp2023-static/icassp2023-static.pgf}
    \vspace*{-0.6cm}
    \caption[]{Mean of 500 post-convergence frames of median NPM of 30 Monte-Carlo runs for fixed \(\gamma_i\) on the interval \([0.0, 0.04]\) and \(K \in \{1,2,10\}\). Results for ideal algorithm, and adaptive \(\gamma_i\) included for comparison.}
    \label{fig:simulations:avgNPMgamma}
\end{figure}
\subsection[]{Communication cost}
\label{sec:transcost}
To describe the communication cost within the network, we count the number of variables transmitted per time frame \(n\).
The most useful measure in the case of this work is to analyze the cost per node.
We compare the following communication schemes within the context of the algorithm \cite{blochbergerDBSI} this paper proposes the extension to:
\begin{itemize}
    \itemsep-0.2em
    \item[(1)] A fully-connected network, where node \(i\) communicates \(\|\bar{\h}_i\|^2\) to all other nodes \(\{j \in \Nset_i = \Mset\}\) directly, i.e., the neighborhood is the full network.
    \item[(2)] The node \(i\) communicates \(\|\bar{\h}_i\|^2\) only to neighboring nodes \(\{j \in \Nset_i \subset \Mset\}\) and applies \(K\) distributed averaging iterations.
\end{itemize}
\autoref{tab:transcost:table} compares the number of transmit and receive operations node \(i\) has to apply in order to compute \eqref{eq:online_admm:consensus_update}.
Further, it shows the complexity of the local problem at node \(i\), assuming it uses all information that it receives.
The advantage of the fully distributed algorithm becomes clear when \(M \gg N_i\), as the additional communication and local complexity depends on neighborhood size \(N_i\) and not on network size \(M\).
\vspace*{-0.6em}
\renewcommand{\arraystretch}{1.2}
\begin{table}[h]
    \centering
    \begin{tabular}{ |l|l|l|l| }
        \hline
        & Trans. Ops. & Rec. Ops. & Local Compl. \\
        \hline\hline
        (1) Fully con. & \(M-1\) & \(M-1\) & \(\mathcal{O}(M^2 L^2)\) \\
        \hline
        (2) Neighborh. & \(N_i K\) & \(N_i K\) & \(\mathcal{O}(N_i^2 L^2)\)\\ 
        \hline
    \end{tabular}
    \caption[]{Communication cost per node and frame \(n\) for computation of \eqref{eq:online_admm:consensus_update}.}
    \label{tab:transcost:table}
\end{table}
\renewcommand{\arraystretch}{1.0}

\subsection[]{Simulations}
To evaluate the effectiveness of the proposed extension, simulations were run, where we define the error measure as the normalized projection misalignment (e.g, \cite{huangAdaptiveMultichannelLeast2002,huangClassFrequencydomainAdaptive2003,habetsOnlineQuasiNewtonAlgorithm2010})
\begin{equation}
        \text{NPM}(n) = 20\,\log_{10} \left(\left\| \hat{\h}^{(n)} - \frac{\h^\T \hat{\h}^{(n)}}{\h^\T \h}\hat{\h}^{(n)} \right\| / \left\|\hat{\h}^{(n)}\right\|\right),
\end{equation}
commonly used in BSI to compare the estimated \(\hat{\h}^{(n)}\) and true \(\h\).

The first simulation setup is a network with \(M=5\) nodes arranged in a ring topology, each node having \(N_i=3\) neighbors (including node \(i\)).
The input signal is zero-mean white Gaussian noise (WGN), the impulse responses of length \(L=16\) are drawn from a normal distribution, and at each channel, independent WGN is added with \(\text{SNR}=10\,\text{dB}\).
The norms of the impulse responses are scaled to random values drawn from the uniform distribution \(\mathcal{U}_{[0.5,2.0]}\).
A comparison is made between the following cases:
\begin{itemize}
    \itemsep-0.2em
    \item[(a)] \(\|\bar{\h}_i\|^2\) of all \(i \in \Mset\) is available for global norm computation (``ideal'' in \autoref{fig:simulations:NPMtime} \& \autoref{fig:simulations:avgNPMgamma}),
    \item[(b)] Inter-neighborhood communication with fixed \(\gamma_i \in [0.0, 0.4]\) and \(K \in \{1,2,10\}\),
    \item[(c)] Inter-neighborhood communication with adaptive \(\gamma_i\) \eqref{eq:adaptivenormest:adaptivegamma} and \(K=1\).
\end{itemize}
\begin{algorithm}[t]
    \caption{ADMM BSI with distributed-averaging-based adaptive estimation of norm values}\label{alg:davg_norm_est}
    \(\W \gets\) \eqref{eq:adaptivenormest:fdlaminprob}\;
    \(\eta_i^{(n-1)} \gets 1, \forall i \in \Mset\)\;
    \For(){\(n=0\dots\)}
    {
        \For(){\(i \in \Mset\)}
        {
            \emph{The steps before are as introduced in }\cite{blochbergerDBSI}\\
            \dotfill\\
            \(\eta_i^{(n)} \gets \sum_{j \in \Nset_i} W_{ij} \|\bar{\h}_i^{(n)}\|^2\)\;
            \(\gamma_i^{(n)} \gets \min \left\lbrace \frac{| \eta_i^{(n)} - \eta_i^{(n-1)} |}{\eta_i^{(n-1)}},\,1\right\rbrace\)\;
            \eIf(){\(n = 0\)}
            {
                \(\phi_i^{(0)}(0) \gets \|\bar{\h}_i^{(0)}\|^2\)\;
            }
            {
                \(\phi_i^{(n)}(0) \gets \gamma_i^{(n)} \eta_i^{(n)} + (1-\gamma_i^{(n)}) \phi_i^{(n-1)}(K)\)\;
            }
            \For(){\(k=0,\dots,K\)}
            {
                Transmit \(\phi_i^{(n)}(k)\) to nodes  \(j \in \Nset_i\)\;
                \(\phi_i^{(n)}({k+1}) \gets \sum_{j \in \Nset_i} W_{ij} \phi_j^{(n)}({k})\)\;
            }
            \(\hat{\h_i}^{(n)} \gets \frac{\bar{\h}_i^{(n)}}{\sqrt{M \phi_i^{(n)}(K)}}\)\;
            \dotfill\\
            \emph{The steps after are as introduced in }\cite{blochbergerDBSI}\\
        }
    }
\end{algorithm}
\autoref{fig:simulations:NPMtime} shows the median of 30 Monte-Carlo runs.
We can observe that when fixing \(\gamma_i\), a tradeoff between convergence speed and steady-state error arises at low iteration counts (cf. also \autoref{fig:simulations:avgNPMgamma}).
With \(K=10\), the performance comes close to the optimal case at the cost of additional communication between nodes.
This in contrast to the case with adaptive \(\gamma_i\) where even at \(K=1\), i.e., no additional communication for the recursive estimation scheme, convergence speed and steady-state error are close to the optimal case.
